MCMC convergence diagnostics

### 2023-07-18 src/docs_paper/appendix.Rmd
Add MCMC output numerics to manuscript and appendix, plus traceplots for worst parameters #55

We assessed MCMC convergence and suitability using a range of graphical and numerical tests.
All potential scale reduction factor $\hat R$ statistics [@vehtari2021rank] were below 1.05 and therefore acceptable (Figure \ref{fig:rhat}).
However, even thinning by a factor of 20, samples were not obtained very efficiently, resulting in the majority of effective sample size (ESS) ratios being below 0.5, with some as low as 0.1 (Figure \ref{fig:ratio}).
As a result, the number of obtained ESS varied substantially by parameter (Figure \ref{fig:ess}).
There were no divergent transitions.
Could add energy plot from @betancourt2017conceptual here.

@@

@@

```{r rho_a, fig.cap="In contrast to the AR1 hyperparameters, the BYM2 hyperparameter posteriors are much less correlated."}
knitr::include_graphics("depends/alpha_x.png")
```

>>>>

```{r}
mcmc_out <- readRDS("depends/mcmc-out.rds")
```

We assessed MCMC convergence and suitability using a range of graphical and numerical tests.
The largest scale reduction factor $\hat R$ was `r signif(mcmc_out$max_rhat, 3)` (Figure \ref{fig:rhat}).
As such, all values of $\hat R < 1.05$.
Even thinning by a factor of 20, samples were not obtained very efficiently, resulting in the majority of effective sample size (ESS) ratios being below 0.5, with some as low as 0.1 (Figure \ref{fig:ratio}).
As a result, the number of obtained ESS varied substantially by parameter (Figure \ref{fig:ess}): the minimum was `r round(mcmc_out$ess_min, 1)`, 2.5% quantile was `r round(mcmc_out$ess_lower, 1)`, median was `r round(mcmc_out$ess_median, 1)`, 97.5% quantile was `r round(mcmc_out$ess_upper, 1)` and maximum was `r round(mcmc_out$ess_max, 1)`.
Traceplots for the parameters with the lowest ESS (`log_sigma_alpha_xs`) and highest $\hat R$ (the 10th index of `ui_lambda_x`) are shown in Figure \ref{fig:worst_trace}.
There were no divergent transitions.
<!-- Could add energy plot from @betancourt2017conceptual here. -->

@@

```{r worst_trace, fig.cap="Traceplots of the parameters with the lowest ESS and highest scale reduction factor."}
knitr::include_graphics("depends/worst-trace.png")
```

@@

```{r alpha_x, fig.cap="In contrast to the AR1 hyperparameters, the BYM2 hyperparameter posteriors are much less correlated."}
knitr::include_graphics("depends/alpha_x.png")
```

### 2023-07-18 src/docs_paper/paper.Rmd
Add MCMC output numerics to manuscript and appendix, plus traceplots for worst parameters #55

\begin{table}[]
\small
\begin{tabularx}{\textwidth}{p{0.15\linewidth}p{0.15\linewidth}p{0.6\linewidth}}
\toprule
Method & Software & Details \\
\midrule
TMB & \texttt{TMB} & $1000$ samples \\
PCA-AGHQ & \texttt{aghq} & $k = 3, s = 8$ (see Section \ref{sec:pca-use}), $1000$ samples \\
NUTS & \texttt{tmbstan} & $4$ chains of $20000$ iterations, with the first $10000$ iterations of each chain discarded as warmup, thinned by a factor of $20$. Default NUTS tuning parameters \citep{hoffman2014no}. \\
\bottomrule
\end{tabularx}
\caption{A summary of settings used for each inferential method.}
\label{tab:inference-methods}
\small
\end{table}

@@

We increased the chain lengths until all NUTS diagnostics were acceptable.
This required chains of length 20,000, which we thinned by a factor of 20 for ease-of-storage.
All potential scale reduction factors [@gelman1992inference; @vehtari2021rank] were $\hat R < 1.05$ (Figure S7).
For full details see Appendix S2.
Though inaccuracies remain possible, we considered the NUTS results to be a gold-standard.

@@

Overlaying the resulting $3^8 = 6561$ PCA-AGHQ nodes onto the hyperparameter marginal posteriors obtained using NUTS, we found approximately 12 of the 24 hyperparameters had well covered marginals (Figure \ref{fig:node-positions}).
Though 12 is an improvement on the 8 that would be naively achieved using a dense grid, there remained many hyperparameters poorly covered.
Coverage was associated with marginal standard deviation (Figure S4), which varied particularly according to the hyperparameter scale.
All constrained hyperparameters $\theta$ were transformed to the real line, using either a log ($\theta > 0$) or logit ($\theta \in [0, 1]$) transformation.
As a result, marginal standard deviations for log transformed hyperparameters were systematically smaller than those which were logit transformed (Figure S5).

@@

To assess the informativeness of the data we compared the prior variance $\sigma_\text{prior}^2(\psi)$ to the posterior variance $\sigma_\text{posterior}^2(\psi)$ via the posterior contraction $c(\psi) = 1 - (\sigma_\text{posterior}^2(\psi) / \sigma_\text{prior}^2(\psi))$ [@schad2021toward], where $\psi$ is a model parameter.
We found that (Figure \ref{fig:contraction})) something something.
For greater interpretability, facet parameters in this plot according to model component.

@@

Let $\Psi = \{\bpsi_i\}_{i = 1}^n$ and $\Psi = \{\bpsi_i\}_{i = 1}^n$ be two sets of joint posterior samples, and $k$ be a kernel.
The maximum mean discrepancy [MMD; @gretton2006kernel] can be empirically estimated by
\begin{equation*}
\text{MMD}(\Psi, \Psi) = \sqrt{\frac{1}{n^2} \sum_{i, j = 1}^n k(\bpsi_i, \bpsi_j) - \frac{2}{n^2} \sum_{i, j = 1}^n k(\bpsi_i, \bpsi_j) + \frac{1}{n^2} \sum_{i, j = 1}^n k(\bpsi_i, \bpsi_j)}.
\end{equation*}
We set $k(\bpsi_i, \bpsi_j) = \exp(-\sigma \lVert \bpsi_i - \bpsi_j \rVert^2)$ with $\sigma$ estimated from data using the `kernlab` \textsc{R} package [@karatzoglou2019package].
As compared with NUTS, the MMD from PCA-AGHQ (0.071) was 11% smaller than that of TMB (0.080).

@@

```{r exceedance, fig.cap="Though PCA-AGHQ does perform slightly better, both approximate inference methods are meaningfully inaccurate as compared with NUTS for estimating exceedance probabilities. For the second 90 target the inaccuracy varies substantially by sex."}
knitr::include_graphics("depends/exceedance.png")
```

>>>>

\begin{table}[]
\small
\begin{tabularx}{\textwidth}{p{0.15\linewidth}p{0.15\linewidth}p{0.6\linewidth}}
\toprule
Method & Software & Details \\
\midrule
TMB & \texttt{TMB} & $1000$ samples \\
PCA-AGHQ & \texttt{aghq} & $k = 3, s = 8$ (see Section \ref{sec:pca-use}), $1000$ samples \\
NUTS & \texttt{tmbstan} & $4$ chains of $100,000$ iterations, with the first $50,000$ iterations of each chain discarded as warmup, thinned by a factor of $40$, to give a total of $5000$ samples kept. Default NUTS tuning parameters \citep{hoffman2014no}. \\
\bottomrule
\end{tabularx}
\caption{A summary of settings used for each inferential method.}
\label{tab:inference-methods}
\small
\end{table}

@@

```{r}
mcmc_out <- readRDS("depends/mcmc-out.rds")
# signif(mcmc_out$max_rhat, 4)
```

Obtaining acceptable NUTS diagnostics required four chains run in parallel for 100,000 iterations, thinned by a factor of 20 for ease-of-storage.
The largest potential scale reduction factor [@gelman1992inference; @vehtari2021rank] was $\hat R = 1.021$ (Figure S7), and there were no divergent transitions.
We considered the NUTS results a gold-standard, though inaccuracies remain possible.
For full details see Appendix S2.

@@

Overlaying the resulting $3^8 = 6561$ PCA-AGHQ nodes onto the hyperparameter marginal posteriors obtained using NUTS, we found approximately 12 of the 24 hyperparameters had well covered marginals (Figure \ref{fig:node-positions}).
Though 12 improves on the 8 naively achievable using a dense grid, there remained many hyperparameters poorly covered.
Coverage was associated with marginal standard deviation (Figure S4), which varied particularly according to the hyperparameter scale.
All constrained hyperparameters $\theta$ were transformed to the real line, using either a log ($\theta > 0$) or logit ($\theta \in [0, 1]$) transformation.
As a result, marginal standard deviations for log transformed hyperparameters were systematically smaller than those which were logit transformed (Figure S5).

@@

To assess the informativeness of the data we compared the prior variance $\sigma_\text{prior}^2(\psi)$ to the posterior variance $\sigma_\text{posterior}^2(\psi)$ via the posterior contraction [@schad2021toward]
\begin{equation}
c(\psi) = 1 - (\sigma_\text{posterior}^2(\psi) / \sigma_\text{prior}^2(\psi)),
\end{equation}
where $\psi$ is a model parameter.
We found that (Figure \ref{fig:contraction})) something something.
For greater interpretability, facet parameters in this plot according to model component.

@@

Let $\Psi^{1} = \{\bpsi^1_i\}_{i = 1}^n$ and $\Psi^2 = \{\bpsi^2_i\}_{i = 1}^n$ be two sets of joint posterior samples, and $k$ be a kernel.
The maximum mean discrepancy [MMD; @gretton2006kernel] can be empirically estimated by
\begin{equation*}
\text{MMD}(\Psi^1, \Psi^2) = \sqrt{\frac{1}{n^2} \sum_{i, j = 1}^n k(\bpsi^1_i, \bpsi^1_j) - \frac{2}{n^2} \sum_{i, j = 1}^n k(\bpsi_i^1, \bpsi_j^2) + \frac{1}{n^2} \sum_{i, j = 1}^n k(\bpsi^2_i, \bpsi^2_j)}.
\end{equation*}
We set $k(\bpsi^1, \bpsi^2) = \exp(-\sigma \lVert \bpsi^1 - \bpsi^2 \rVert^2)$ with $\sigma$ estimated from data using the `kernlab` \textsc{R} package [@karatzoglou2019package].
As compared with NUTS, the MMD from PCA-AGHQ (0.071) was 11% smaller than that of TMB (0.080).

@@

```{r exceedance, fig.cap="Though PCA-AGHQ was marginally better, both approximate inference methods were meaningfully inaccurate as compared with NUTS for estimating exceedance probabilities. For the second 90 target the inaccuracy varied substantially by sex."}
knitr::include_graphics("depends/exceedance.png")
```

### 2023-07-18 src/naomi-simple_mcmc/mcmc-convergence.Rmd
Add MCMC output numerics to manuscript and appendix, plus traceplots for worst parameters #55

```{r}
out <- readRDS("depends/out.rds")
mcmc <- out$mcmc$stanfit
```

@@

bayesplot::color_scheme_set("viridisA")
ggplot2::theme_set(theme_minimal())
```

@@

We are looking for values of $\hat R$ less than 1.1 here.

@@

(big_rhats <- rhats[rhats > 1.1])
length(big_rhats) / length(rhats)
```

@@

ggsave("ratio.png", h = 3, w = 6.25)
```

@@

ggsave("ess.png", h = 3, w = 6.25)
```

@@

@@

```{r}
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_a", "logit_phi_alpha_a"), diag_fun = "hist", off_diag_fun = "hex")
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_as", "logit_phi_alpha_as"), diag_fun = "hist", off_diag_fun = "hex")
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_as", "logit_phi_rho_as"), diag_fun = "hist", off_diag_fun = "hex")
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_a", "logit_phi_rho_a"), diag_fun = "hist", off_diag_fun = "hex")

ggsave("rho_a.png", h = 4, w = 6.25)

@@

```{r}
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_x", "logit_phi_rho_x"), diag_fun = "hist", off_diag_fun = "hex")
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_xs", "logit_phi_rho_xs"), diag_fun = "hist", off_diag_fun = "hex")
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_xs", "logit_phi_alpha_xs"), diag_fun = "hist", off_diag_fun = "hex")
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_x", "logit_phi_alpha_x"), diag_fun = "hist", off_diag_fun = "hex")

ggsave("alpha_x.png", h = 4, w = 6.25)

@@

neighbours_pairs_plot <- function(par, i) {
  neighbour_pars <- paste0(par, "[", c(i, nb[[i]]), "]")
  bayesplot::mcmc_pairs(mcmc, pars = neighbour_pars, diag_fun = "hist", off_diag_fun = "hex")
}

# area_merged %>%
#   filter(area_level == max(area_level)) %>%
#   print(n = Inf)
```

>>>>

```{r}
out <- readRDS("depends/out.rds")
mcmc <- out$mcmc$stanfit

depends <- yaml::read_yaml("orderly.yml")$depends

dependency_details <- function(i) {
  report_name <- names(depends[[i]])
  print(paste0("Inference results obtained from ", report_name, " with the query ", depends[[i]][[report_name]]$id))
  report_id <- orderly::orderly_search(query = depends[[i]][[report_name]]$id, report_name)
  print(paste0("Obtained report had ID ", report_id, " and was run with the following parameters:"))
  print(orderly::orderly_info(report_id, report_name)$parameters)
}

dependency_details(1)
```

@@

color_scheme_set("viridis")
ggplot2::theme_set(theme_minimal())
```

@@

We are looking for values of $\hat R$ less than 1.05 here.

@@

(big_rhats <- rhats[rhats > 1.05])
length(big_rhats) / length(rhats)

(max_rhat <- max(rhats))
```

@@

ggsave("ratio.png", h = 3, w = 6.25)

(average_ess_ratio <- mean(ratios))
```

@@

ggsave("ess.png", h = 3, w = 6.25)

(ess_min <- min(mcmc_summary[, "n_eff"]))
(ess_lower <- quantile(mcmc_summary[, "n_eff"], 0.025))
(ess_median <- quantile(mcmc_summary[, "n_eff"], 0.50))
(ess_upper <- quantile(mcmc_summary[, "n_eff"], 0.975))
(ess_max <- max(mcmc_summary[, "n_eff"]))
```

Save outputs for use in manuscript:

```{r}
out <- list(
  "max_rhat" = max_rhat,
  "average_ess_ratio" = average_ess_ratio, 
  "ess_min" = ess_min,
  "ess_lower" = ess_lower,
  "ess_median" = ess_median,
  "ess_upper" = ess_upper,
  "ess_max" = ess_max
)

saveRDS(out, "out.rds")
```

@@

The parameters with the worst ESS and the worst $\hat R$:

```{r}
(plot <- bayesplot::mcmc_trace(mcmc, pars = c(names(which.min(mcmc_summary[, "n_eff"])), names(which.max(rhats)))))

ggsave("worst-trace.png", plot, h = 4, w = 6.25)
```

@@

```{r}
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_a", "logit_phi_alpha_a"), diag_fun = "hist", off_diag_fun = "hex")
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_as", "logit_phi_alpha_as"), diag_fun = "hist", off_diag_fun = "hex")
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_as", "logit_phi_rho_as"), diag_fun = "hist", off_diag_fun = "hex")
(plot <- bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_a", "logit_phi_rho_a"), diag_fun = "hist", off_diag_fun = "hex"))

ggsave("rho_a.png", plot, h = 4, w = 6.25)

@@

```{r}
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_x", "logit_phi_rho_x"), diag_fun = "hist", off_diag_fun = "hex")
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_xs", "logit_phi_rho_xs"), diag_fun = "hist", off_diag_fun = "hex")
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_xs", "logit_phi_alpha_xs"), diag_fun = "hist", off_diag_fun = "hex")
(plot <- bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_x", "logit_phi_alpha_x"), diag_fun = "hist", off_diag_fun = "hex"))

ggsave("alpha_x.png", plot, h = 4, w = 6.25)

@@

neighbours_pairs_plot <- function(par, i) {
  neighbour_pars <- paste0(par, "[", c(i, nb[[i]]), "]")
  bayesplot::mcmc_pairs(mcmc, pars = neighbour_pars, diag_fun = "hist", off_diag_fun = "hex")
}
```
